Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-14 notebooks/expected_and_actual_counts_ratio.Rmd
Pull updated CSV file and modify amino acid analyses to use updated file

#install.packages("data.table")
#install.packages("ggplot2")
#install.packages("ggforce")
#install.packages("RColorBrewer")
#install.packages("stringr")
#install.packages("gtools")

library(data.table)
library(ggplot2)
library(ggforce)
library(RColorBrewer)
library(stringr)
library(gtools)

@@

```{r read_data}
# Read data using access to the Fred Hutch server.
SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)

# Read data from github.com.
#SARS_mutation <- fread("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
```

@@

With the data set, both nucleotide and amino acid mutations can be compared between clades. However, there is no column identifying the amino acid mutation (i.e., starting amino acid, codon position along the genome and mutated amino acid). Create a column with this information using the nucleotide site, codon position (i.e., 1, 2, or 3), clade founder amino acid and mutant amino acid columns.

```{r aa_mutation}
if(option=='aa'){
  # Create a list with the starting nucleotide site for an amino acid for all rows in the data set.
  starting_nt <- SARS_mutation$nt_site-as.numeric(str_sub(SARS_mutation$codon_position,1,1))+1

  # Create a column with amino acid mutation (e.g., M.266.268.L).
  SARS_mutation$aa_mutation <- paste(str_sub(SARS_mutation$clade_founder_aa,1,1), starting_nt, starting_nt+2, str_sub(SARS_mutation$mutant_aa,1,1), sep=".")
}

# NEED TO SUM TOGETHER DUPLICATE AMINO ACIDS
# consider entire gene column
# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
```

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)

# Remove columns that were used to filter the data.
filtered_SARS_mutation$subset <- NULL
filtered_SARS_mutation$exclude <- NULL
```

@@

# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,-1,-1), sep="")
(table(aa_type)/length(aa_type))*100
```

@@

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)

stacked_aa <- filtered_SARS_mutation
stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder, ncol=7)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

With such a large data set, it could take hours to calculate the p-values for all clade comparisons. To decrease the amount of time it takes to conduct analyses, this notebook will only compare two clades at a time and the script can be run in parallel to generate all comparisons at once. Therefore, filter the data to only include mutation data from the two clades being compared.

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

@@

```{r restructure_data}
# Create two data frames with data for clade 1 and clade 2.
clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),]
clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),]

@@

# Determine if any of the columns from 'clade1' and 'clade2' are the same.
col_names <- colnames(filtered_clades)[colnames(filtered_clades) %notin% merge_column]
is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[, get(paste(c[1], 1, sep="_"))], comparisons[, get(paste(c[1], 2, sep="_"))])}))
col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]

@@

```{r volcano_corrected, echo=FALSE}
# After dropping P = 1, plot the corrected p-values.
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(comparison_p_values[,str_sub(get(paste(option, "mutation", sep="_")),1,1)])))

volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(paste(option, "mutation", sep="_")), color=str_sub(get(paste(option, "mutation", sep="_")),1,1)))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()
```

@@

```{r verify_analyses, echo=FALSE}
# Plot the corrected p-values for synonymous and non-synonymous mutations.
verify_analyses <- comparison_p_values
if("synonymous" %notin% colnames(verify_analyses)){
  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="TRUE" & verify_analyses$synonymous_2=="TRUE", "TRUE", "FALSE")
}
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")

>>>>

#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("ggforce")
#install.packages("RColorBrewer")
#install.packages("stringr")
#install.packages("gtools")

library(dplyr)
library(ggplot2)
library(ggforce)
library(RColorBrewer)
library(stringr)
library(gtools)

@@

```{r read_data}
# Read data using access to the Fred Hutch server.
SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)

# Read data from github.com.
#SARS_mutation <- read.csv("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
```

@@

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="False" & expected_count>=min_expected)
```

```{r aa_mutation}
if(option=='aa'){
  aa_group <- filtered_SARS_mutation
  filtered_SARS_mutation <- as.data.frame(aa_group %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
}else{
  nt_group <- filtered_SARS_mutation
  filtered_SARS_mutation <- nt_group[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","actual_count","expected_count")]
}

# consider entire gene column
# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
```

@@

# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,1,1), sep="")
aa_type <- paste(filtered_SARS_mutation$clade_founder_aa, ".", filtered_SARS_mutation$mutant_aa, sep="")
(table(aa_type)/length(aa_type))*100
```

@@

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(79)

stacked_aa <- filtered_SARS_mutation
#stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
#stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder_aa, ncol=7)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

@@

```{r restructure_data}
# Create two data frames with data for clade 1 and clade 2.
clade1 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[1]),]
clade2 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[2]),]

@@

# Determine if any of the columns from 'clade1' and 'clade2' are the same.
col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% merge_column]
is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]

@@

```{r volcano_corrected, echo=FALSE}
# After dropping P = 1, plot the corrected p-values.
color <- colorRampPalette(brewer.pal(8,"Set1"))(comparison_p_values[,paste("clade_founder", option, sep="_")] %>% unique() %>% length())

volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=get(paste("clade_founder", option, sep="_"))))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()
```

@@

```{r verify_analyses, echo=FALSE}
# Plot the corrected p-values for synonymous and non-synonymous mutations.
verify_analyses <- comparison_p_values
if("synonymous" %notin% colnames(verify_analyses)){
  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="True" & verify_analyses$synonymous_2=="True", "True", "False")
}
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
